---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Project Business Economic and Financial Data

## Sales of DimSum Records, Asian-food restaurant in Medellin, Colombia

2024/2025

Authors: Daniel Gutierrez & Fabio Pimentel

```{r}
# Required Packages--------------------
rm(list = ls())
library(readxl)
library(ggplot2)
library(GGally)
library(dplyr)
library(lubridate)
library(corrplot)
library(feasts)
library(tsibble)
library(forecast)
library(tidyr)
library(ggthemes)
library(car)
library(DIMORA)
library(tseries)
library(lmtest)
library(prophet)
```

### Import Data

```{r}
#setwd('/Users/fabiopimentel/Documents/Padua/clases/segundo año primer semestre/BEF data/proyecto/time_series_padova-main')

# target variable
sales <- read_excel("data/sales/sales_dimsum_31102024.xlsx")

sales[is.na(sales)] <- 0 # set to zero na values

```

### Creating variables

```{r}
# economic variables
eco_growth <- read_excel("data/macroeconomic/economic_activity.xlsx")
fx <- read_excel("data/macroeconomic/fx.xlsx") #Foreign exchange is the conversion of one currency into another
inflation <- read_excel("data/macroeconomic/inflation.xlsx")
unemployment <- read_excel("data/macroeconomic/unemployment.xlsx")
```

```{r}
# other variables
google_trends <- read_excel("data/other/google_trends_restaurantes.xlsx")
rain <- read_excel("data/other/rain_proxy.xlsx")
temp <- read_excel("data/other/temperature_data.xlsx")
temp[is.na(temp)] <- 0
rain[is.na(rain)] <- 0
plot(temp$tavg) # no zeros in temp : OK
plot(temp$tmedian) # no zeros in temp : OK- looks better than mean
```

### Explore data structure

```{r}
str(sales)
```

```{r}
str(eco_growth)
```

```{r}
str(fx) # Foreign exchange is the conversion of one currency into another
```

```{r}
str(inflation)
```

```{r}
str(unemployment)
```

```{r}
str(google_trends)
```

```{r}
str(rain)
```

```{r}
str(temp) # this has NaNs, must fill somehow
```

### Sales

```{r}
# create time variables
plot(sales$sales_cop)
```

### Sales Monthly

```{r}
# sales
## sales monthly
df_sales_m <- sales %>%
  mutate(month = floor_date(date, "month")) %>% # Extract month
  group_by(month) %>%
  summarise(sales_m = sum(sales_cop), bar_m = sum(bar), food_m = sum(food)
            )     # Summing values

head(df_sales_m)
```

### Sales Weekly

```{r}
## sales weekly
df_sales_w <- sales %>%
  mutate(week = floor_date(date, "week")) %>% # Extract month
  group_by(week) %>%
  summarise(sales_w = sum(sales_cop), bar_w = sum(bar), food_w = sum(food))     # Summing values

head(df_sales_w)
```

### FX

```{r}
# fx
df_fx_m <- fx %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(fx_m = mean(fx))

df_fx_w <- fx %>%
  mutate(week = floor_date(date, "week")) %>%
  group_by(week) %>%
  summarise(fx_w = mean(fx))

head(df_fx_m)
head(df_fx_w)
```

### Google Trends

```{r}
# google trends

# montly
df_google_m <- google_trends %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(google_m = mean(google_trends))


# weekly
df_google_w <- google_trends %>%
  mutate(week = floor_date(date, "week")) %>%
  group_by(week) %>%
  summarise(google_w = mean(google_trends))

head(df_google_m)
head(df_google_w)
```

### Rain

```{r}
## rain
df_rain_g = rain %>%
  group_by(date, region) %>%
  summarise(rain_sum=sum(contribution_m3s))

df_rain_g  <- df_rain_g[df_rain_g$region=="ANTIOQUIA",]

head(df_rain_g)

# montly
df_rain_m <- df_rain_g %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(rain_m = sum(rain_sum))


# weekly
df_rain_w <- df_rain_g %>%
  mutate(week = floor_date(date, "week")) %>%
  group_by(week) %>%
  summarise(rain_w = sum(rain_sum))

head(df_rain_m)
head(df_rain_w)
```

### Temperature

```{r}
# temperature
# montly
df_temp_m <- temp %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(temp_m = mean(tavg), prcp_m = sum(prcp))


# weekly
df_temp_w <- temp %>%
  mutate(week = floor_date(date, "week")) %>%
  group_by(week) %>%
  summarise(temp_w = mean(tavg), prcp_w = sum(prcp))

head(df_temp_m)
head(df_temp_w)
```

## Merging Data Frames

#### Daily Data

```{r}
## daily data----------
#sales, rain, fx are the only ones daily
df_merged_d <- merge(sales, df_rain_g, by = "date", all = FALSE) # Inner join
df_merged_d <- merge(df_merged_d, fx, by = "date", all = FALSE) # Inner join
df_merged_d <- merge(df_merged_d, temp, by = "date", all = FALSE) # Inner join

head(df_merged_d)

```

#### Weekly Data

```{r}
### weekly data----------
df_merged_w <- merge(df_sales_w, df_rain_w, by="week", all=F)
df_merged_w <- merge(df_merged_w, df_google_w, by="week", all=F)
df_merged_w <- merge(df_merged_w, df_fx_w, by="week", all=F)
df_merged_w <- merge(df_merged_w, df_temp_w, by="week", all=F)

head(df_merged_w)
```

#### Monthly Data

```{r}
### monthly data----------
# change colnames
names(eco_growth) <- c("month", "ise")
names(inflation) <- c("month", "inflation")
names(unemployment) <- c("month", "unemployment") 

df_merged_m <- merge(df_sales_m, df_rain_m, by="month", all=F)
nrow(df_merged_m)

df_merged_m <- merge(df_merged_m, df_fx_m, by="month", all=F)
nrow(df_merged_m)

df_merged_m <- merge(df_merged_m, df_google_m, by="month", all=F)
nrow(df_merged_m)

df_merged_m <- merge(df_merged_m, eco_growth, by="month", all=F) # only has until aug 2024
nrow(df_merged_m)

df_merged_m <- merge(df_merged_m, inflation, by="month", all=F)
nrow(df_merged_m)

df_merged_m <- merge(df_merged_m, unemployment, by="month", all=F)
nrow(df_merged_m)

df_merged_m <- merge(df_merged_m, df_temp_m, by="month", all=F)
nrow(df_merged_m)
```

### EDA

```{r}
objects_to_keep <- c("df_merged_d", "df_merged_w", "df_merged_m")
# Remove all objects except those specified
rm(list = setdiff(ls(), objects_to_keep))
```

#### Daily Sales

```{r}
# sales daily
ggplot(
  df_merged_d, 
  aes(x=date, y=sales_cop)
  ) + geom_line() + ggtitle("Daily Sales of Restaurant")
```

#### Weekly sales

```{r}
# sales weekly
ggplot(df_merged_w, aes(x=week, y=sales_w)) +
  geom_line() + ggtitle("Weekly Sales of Restaurant")
```

#### Monthly sales

```{r}
# sales montly
ggplot(df_merged_m, aes(x=month, y=sales_m)) +
  geom_line() + ggtitle("Monthly Sales of Restaurant")
```

#### Stacked bar plots

We want to move to a stacked bar chart when we care about the relative
decomposition of each primary bar based on the levels of a second
categorical variable. Each bar is now comprised of a number of sub-bars,
each one corresponding with a level of a secondary categorical variable.
The total length of each stacked bar is the same as before, but now we
can see how the secondary groups contributed to that total.

One important consideration in building a stacked bar chart is to decide
which of the two categorical variables will be the primary variable
(dictating major axis positions and overall bar lengths) and which will
be the secondary (dictating how each primary bar will be subdivided).
The most 'important' variable should be the primary; use domain
knowledge and the specific type of categorical variables to make a
decision on how to assign your categorical variables

```{r}
#Monthly
# Reshape the data to a long format
df_sales_m_long <- df_merged_m %>%
  pivot_longer(cols = c(bar_m, food_m), names_to = "Category", values_to = "Value")

# Create the stacked bar plot
ggplot(df_sales_m_long, aes(x = month, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  ggtitle("Monthly Sales of Restaurant") +
  labs(y = "Sales", x = "Month", fill = "Category") +
  theme_minimal()
```

```{r}
# Weekly
# Reshape the data to a long format
df_sales_w_long <- df_merged
```



```{r}
df_sales_w_long <- df_merged_w %>%
  pivot_longer(cols = c(bar_w, food_w), names_to = "Category", values_to = "Value")

# Create the stacked bar plot
ggplot(df_sales_w_long, aes(x = week, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  ggtitle("Weekly Sales of Restaurant") +
  labs(y = "Sales", x = "Week", fill = "Category") +
  theme_minimal()
```

#### Seasonal plots

```{r}
# Seasonal plots
df_sales_w_filtered <- df_merged_w %>%
  filter(week >= ymd("2021-12-31"))


tseries_w <- ts(df_sales_w_filtered$sales_w , start = c(2022, 1), frequency = 52)
head(tseries_w)
seasonplot(tseries_w, col = rainbow(3), year.labels = TRUE, main = "Seasonal Plot")
text(x = 1, y = max(tseries_w) - 1.5e7, labels = "2024", col = "blue")


```

```{r}
# seasonplot monthly
df_sales_m_filtered <- df_merged_m %>%
  filter(month >= ymd("2021-12-31"))


tseries_m <- ts(df_sales_m_filtered$sales_m , start = c(2022, 1), frequency = 12)
head(tseries_m)
seasonplot(tseries_m, col = rainbow(3), year.labels = TRUE, main = "Seasonal Plot")
text(x = 1, y = max(tseries_m) - 1e6, labels = "2024", col = "blue")

```
## Density
```{r}
# Montly Density
# Select the columns of interest
variables <- c("sales_m", "bar_m", "food_m", "rain_m", "fx_m", "google_m",
               "ise", "inflation", "unemployment", "temp_m", "prcp_m")


# Transform the data to long format for ggplot2
df_long_m <- df_merged_m %>%
  pivot_longer(cols = all_of(variables), names_to = "Variable", values_to = "Value")

# Create the grid of density plots
ggplot(df_long_m, aes(x = Value)) +
  geom_density(fill = "blue", alpha = 0.4) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(title = "Density Plots of Selected Variables",
       x = "Value", y = "Density") +
  theme_minimal()

```

```{r}
# Weekly Density
# Select the columns of interest
variables <- c("sales_w", "bar_w", "food_w", "rain_w", "fx_w", "google_w",
                "temp_w", "prcp_w")



df_long_w <- df_merged_w %>%
  pivot_longer(cols = all_of(variables), names_to = "Variable", values_to = "Value")

# Create the grid of density plots
ggplot(df_long_w, aes(x = Value)) +
  geom_density(fill = "blue", alpha = 0.4) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(title = "Density Plots of Selected Variables",
       x = "Value", y = "Density") +
  theme_minimal()

```


```{r}
# Daily Density

# Select the columns of interest
variables <- c("sales_cop", "bar", "food", "rain_sum", "fx", 
               "tmedian", "prcp")



df_long_d <- df_merged_d %>%
  pivot_longer(cols = all_of(variables), names_to = "Variable", values_to = "Value")

# Create the grid of density plots
ggplot(df_long_d, aes(x = Value)) +
  geom_density(fill = "blue", alpha = 0.4) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(title = "Density Plots of Selected Variables",
       x = "Value", y = "Density") +
  theme_minimal()

```

## Covariates

```{r}
### 3.5.1 economic variables-----------------------
# economic growth
ggplot(df_merged_m, aes(x=month, y=ise)) +
  geom_line() + ggtitle("Monthly activity in Colombia")
# clearly seasonal and trend

# fx
ggplot(df_merged_d, aes(x=date, y=fx)) +
  geom_line() + ggtitle("Daily COP/USD")
# trend but no clear seasonality

# inflation
ggplot(df_merged_m, aes(x=month, y=inflation)) +
  geom_line() + ggtitle("Monthly inflation National")
# business cycles, no tend or seasonality

# unemployment
ggplot(df_merged_m, aes(x=month, y=unemployment)) +
  geom_line() + ggtitle("Montly trailing unemployment Medellin")
# seasonal and trend downwards


### 3.5.2 Other variables

# google trends
ggplot(df_merged_w, aes(x=week, y=google_w)) +
  geom_line() + ggtitle("Weelkly Google trends 'Restaurantes'")
# no clear behaviour, drop in pandemic

# rain
ggplot(df_merged_d, aes(x=date, y=rain_sum)) +
  geom_line() + ggtitle("Daily rain approximated in Antioquia")
# no trend or seasonality clearly

# temperature
ggplot(df_merged_d, aes(x=date, y=tmedian)) +
  geom_line() + ggtitle("Daily Median temperature in Medellin")

# almost stationary

# temperature
ggplot(df_merged_d, aes(x=date, y=tavg)) +
  geom_line() + ggtitle("Daily Average temperature in Medellin")


# this one looks weird, better keep working on median

# precipitation from temp
ggplot(df_merged_d, aes(x=date, y=prcp)) +
  geom_line() + ggtitle("Daily  precipitation in Medellin")
# looks decent

```

### Pairplots

```{r}
df_merged_d <- subset(df_merged_d, select = -region)

# daily
ggpairs(df_merged_d, 
        columns = 2:8)
# sales have correl with fx and rain_sum
# weekly
ggpairs(df_merged_w, 
        columns = 2:9)
# sales have correl with rain, google, fx, temp
# bar has more correl with temp

# montly
ggpairs(df_merged_m, 
        columns = 2:12)

```

### Correlation Matrix

```{r}
# Exclude 'date' column
numeric_df_d <- df_merged_d[, sapply(df_merged_d, is.numeric)]
cor_matrix_d <- cor(numeric_df_d, use = "complete.obs")  # Use only complete rows
cor_matrix_d

numeric_df_w <- df_merged_w[, sapply(df_merged_w, is.numeric)]
cor_matrix_w <- cor(numeric_df_w, use = "complete.obs")  # Use only complete rows
cor_matrix_w

numeric_df_m <- df_merged_m[, sapply(df_merged_m, is.numeric)]
cor_matrix_m <- cor(numeric_df_m, use = "complete.obs")  # Use only complete rows
cor_matrix_m

# Plot the Correlation Matrix
par(mfrow=c(1,1))
corrplot(cor_matrix_d, method = "color", type = "upper", tl.col = "black", tl.srt = 45)
corrplot(cor_matrix_w, method = "color", type = "upper", tl.col = "black", tl.srt = 45)
corrplot(cor_matrix_m, method = "color", type = "upper", tl.col = "black", tl.srt = 45)

```
Rain has stronger correlation than prcp, so we drop prcp to not repeat the same variable from two sources
Also we drop average temperature because median temperature seems more trustworthy

```{r}
# drop prcp beacuse they "are the same"
df_merged_m <- df_merged_m %>% select(-prcp_m)
df_merged_w <- df_merged_w %>% select(-prcp_w)
df_merged_d <- df_merged_d %>% select(-prcp)

# drop avg temp
df_merged_d <- df_merged_d %>% select(-tavg)
colnames(df_merged_d)
```
```{r}
### drop everything not on use
objects_to_keep <- c("df_merged_d", "df_merged_w", "df_merged_m")
# Remove all objects except those specified
rm(list = setdiff(ls(), objects_to_keep))
```


### Variable Transformation

POSIXct and POSIXlt Classes

Times and date-times are represented by the POSIXct or the POSIXlt class
in R. The POSIXct format stores date and time in seconds with the number
of seconds beginning at January 1, 1970, so a POSIXct date-time is
essentially an single value on a timeline. Date-times prior to 1970,
will be negative numbers. The POSIXlt class stores other date and time
information in a list such as hour of day of week, month of year, etc.
The starting year for POSIXlt data is 1900, so 2022 would be stored as
year 122. Months also begin at 0, so January is stored as month 0 and
February as month 1. For both POSIX classes, the timezone can be
classified. While date-times stored as POSIXct and POSIXlt look similar,
when you unclass them with the unclass() function, you can see the
additional information stored within the POSIXlt data.

Date Class

Dates without time can simply be stored as a Date class in R using the
as.Date() function. Both Dates and POXIC classes need to be defined
based on how they formatted. When uploading time series data into R,
date and date-time data is typically uploaded as a character class and
must be converted to date or time class using the as.Date(),
as.POSIXct() or as.POSIXlt() functions.

Monthly

```{r}
# Vars for model
# Month
# Ensure the `month` column is in POSIXct format
df_merged_m$month <- as.POSIXct(df_merged_m$month)

# Create the numeric variable: an evenly increasing number
df_merged_m <- df_merged_m %>%
  arrange(month) %>%  # Ensure data is sorted by month
  mutate(numeric_month = row_number())  # Assign an increasing number

# Create the seasonal variable: the 12 different months as a factor
df_merged_m <- df_merged_m %>%
  mutate(seasonal_month = factor(format(month, "%B"), levels = month.name))  # Month names as ordered factors
```

Weekly

```{r}
# Week
# Ensure the `week` column is in POSIXct format
df_merged_w$week <- as.POSIXct(df_merged_w$week)

# Create the numeric variable: an evenly increasing number
df_merged_w <- df_merged_w %>%
  arrange(week) %>%  # Ensure data is sorted by week
  mutate(numeric_week = row_number())  # Assign an increasing number

# Create the seasonal variable: the 12 different months as a factor
df_merged_w <- df_merged_w %>%
  mutate(seasonal_month = factor(format(week, "%B"), levels = month.name))  # Month names as ordered factors
```

Daily

```{r}
# Day
# Ensure the `day` column is in POSIXct format
df_merged_d$date <- as.POSIXct(df_merged_d$date)

# Create the numeric variable: an evenly increasing number
df_merged_d <- df_merged_d %>%
  arrange(date) %>%  # Ensure data is sorted by day
  mutate(numeric_day = row_number())  # Assign an increasing number
```

```{r}
# Create the seasonal variable: the 12 different months as a factor
df_merged_d <- df_merged_d %>%
  mutate(seasonal_month = factor(format(date, "%B"), levels = month.name))  # Month names as ordered factors

# Create a column indicating the day of the week
df_merged_d <- df_merged_d %>%
  mutate(day_of_week = factor(weekdays(date), levels = c("Monday", "Tuesday", "Wednesday", 
                                                         "Thursday", "Friday", "Saturday", "Sunday")))  # Day of the week as ordered factor
```

### Time Series Objects
Convert sales to time series objects for the use in several models

```{r}
# convert to time series
sales_d_ts <- ts(df_merged_d$sales_cop)
sales_w_ts <- ts(df_merged_w$sales_w)
sales_m_ts <- ts(df_merged_m$sales_m)

par(mfrow=c(1,1))

# Daily
tsdisplay(sales_d_ts)
# is not stationary but has no clear trend
# and seasonality every 7 days

# Weekly
tsdisplay(sales_w_ts)
# not stationary: has trend

# Montly
tsdisplay(sales_m_ts)
# has clear trend, no seasonality

```

### Log Transformation
Some variables are scaled to log, so we can interpret the linear models more easily. The covariates are in different scales so it is easier to interpret percentage changes instead of unit changes.

```{r}
# Monthly
df_merged_m <- df_merged_m %>%
  mutate(across(where(is.numeric) & !all_of(c("unemployment", "inflation")), ~ log(. + 1)))

# Weekly
df_merged_w <- df_merged_w %>%
  mutate(across(where(is.numeric), ~ log(. + 1)))

# Daily
# Weekly
df_merged_d <- df_merged_d %>%
  mutate(across(where(is.numeric), ~ log(. + 1)))

```

### Autocorrelation

```{r}
#par(mfrow=c(1,1))
#tsdisplay(sales_d_ts)
# is not stationary but has no clear trend

plot(sales_d_ts)
acf(sales_d_ts)
pacf(sales_d_ts)
```

When data are seasonal, the autocorrelation will be larger for the
seasonal lags (at multiples of the seasonal period) than for other lags.

```{r}
# Weekly

#tsdisplay(sales_w_ts)
plot(sales_w_ts)
acf(sales_w_ts)
pacf(sales_w_ts)

# not stationary: has trend and seasonality maybe
```

```{r}
# Montly

#tsdisplay(sales_m_ts)
plot(sales_m_ts)
acf(sales_m_ts)
pacf(sales_m_ts)
# has clear trend, no seasonality
```


# Models
In this section we model the time series using various approaches to find the best model for our data.
We use both linear and non linear models going from the simplest to the more "complex" models.


## Helper functions
Functions that help us implement and analyze models faster
```{r}
## Function to create and summarize models------------------
run_model <- function(formula, data, model_name) {
  cat("\nRunning", model_name, "\n")
  model <- lm(formula, data = data)
  print(summary(model))
  par(mfrow = c(2, 2))
  plot(model)
  return(model)
}

# Function to compare models using ANOVA
compare_models <- function(model1, model2, name1, name2) {
  cat("\nComparing Models:", name1, "vs", name2, "\n")
  anova_result <- anova(model1, model2)
  print(anova_result)
  return(anova_result)
}

# Function to add predictions to the dataset
add_predictions <- function(model, data, pred_column) {
  data[[pred_column]] <- predict(model, newdata = data)
  return(data)
}

# Calculate RMSE
# Function to calculate RMSE
calculate_rmse <- function(observed, predicted) {
  rmse <- sqrt(mean((observed - predicted)^2, na.rm = TRUE))
  return(rmse)
}


# function that compares linear models
# Define the function to get R^2 and AIC
get_model_stats <- function(models) {
  # Initialize an empty data frame
  stats <- data.frame(
    Model = character(),
    R2 = numeric(),
    AIC = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through the list of models
  for (i in seq_along(models)) {
    model <- models[[i]]
    model_name <- names(models)[i]
    # Extract R^2 and AIC
    r2 <- summary(model)$r.squared
    aic <- AIC(model)
    # Append to the data frame
    stats <- rbind(stats, data.frame(Model = model_name, R2 = r2, AIC = aic))
  }
  
  return(stats)
}

```


## Linear models

```{r}
# Montly Models
# View Dataframe
head(df_merged_m)

# Model 0: Trend only
ols0 <- run_model(sales_m ~ numeric_month, df_merged_m, "Model 0")
df_merged_m <- add_predictions(ols0, df_merged_m, "predicted_sales0")

# Model 1: Trend + Seasonality
ols1 <- run_model(sales_m ~ numeric_month + seasonal_month, df_merged_m, "Model 1")
df_merged_m <- add_predictions(ols1, df_merged_m, "predicted_sales1")


## Model 2: Backward Stepwise Regression 

# Start with the full model (excluding food and bar)
ols2_full <- lm(
  sales_m ~ numeric_month + seasonal_month + unemployment + ise + fx_m +
    google_m + temp_m + rain_m, 
  data = df_merged_m
)


# Perform backward stepwise regression
ols2_stepwise <- step(
  ols2_full, 
  direction = "backward",
  trace = 1 # Prints the stepwise regression process
)

# Summary of the final stepwise model
summary(ols2_stepwise)

# Add predictions from the final stepwise model
df_merged_m <- add_predictions(ols2_stepwise, df_merged_m, "predicted_sales2")

# Plot Actual vs Predicted Values
ggplot(df_merged_m, aes(x = month)) +
  geom_line(aes(y = exp(sales_m), color = "Actual Sales"), size = 1) +
  geom_line(aes(y = exp(predicted_sales0), color = "Model 0"), linetype = "dashed", size = 1) +
  geom_line(aes(y = exp(predicted_sales1), color = "Model 1"), linetype = "dotted", size = 1) +
  geom_line(aes(y = exp(predicted_sales2), color = "Model 2 Stepwise"), linetype = "dotdash", size = 1) +
  labs(title = "Actual vs Predicted Monthly Sales for All Models",
       x = "Month", y = "Sales", color = "Legend") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Models to compare
models <- list(
  "Model trend" = ols0,
  "Model trend + season" = ols1,
  "Model all covariates step" = ols2_stepwise
)

# Get R^2 and AIC for each model
model_stats <- get_model_stats(models)

# View the results
print(model_stats)


#  RMSE calculation for the original (exponentiated) scale
rmse_stats <- data.frame(
  Model = character(),
  RMSE = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each model
for (i in seq_along(models)) {
  model_name <- names(models)[i]
  predicted_column <- paste0("predicted_sales", i - 1)  # Adjust column name index
  
  # Calculate RMSE on the original scale
  rmse <- calculate_rmse(
    observed = exp(df_merged_m$sales_m),          # Exponentiate actual values
    predicted = exp(df_merged_m[[predicted_column]])  # Exponentiate predicted values
  )
  
  # Append results to the RMSE stats table
  rmse_stats <- rbind(rmse_stats, data.frame(Model = model_name, RMSE = rmse))
}

# View RMSE statistics
print(rmse_stats)

```


```{r}
rmse_ols_m <- rmse_stats$RMSE[3]
rmse_ols_m
```

```{r}
# Weekly Models
head(df_merged_w)
## Clean Data - Drop rows 1-2 because sales are 0 / was not open yet
df_merged_w <- df_merged_w %>% slice(-1, -2)

## Model 0A: Trend only
ols0w <- run_model(sales_w ~ numeric_week, df_merged_w, "Model 0A")
df_merged_w <- add_predictions(ols0w, df_merged_w, "predicted_sales0")

## Model 1A: Trend + Seasonality
ols1w <- run_model(sales_w ~ numeric_week + seasonal_month, df_merged_w, "Model 1A")
df_merged_w <- add_predictions(ols1w, df_merged_w, "predicted_sales1")


## Model 2A: Experimentation


# Start with the full model (excluding food and bar)
ols2_full_w <- lm(
  sales_w ~ numeric_week + seasonal_month + fx_w +
    google_w + temp_w + rain_w, 
  data = df_merged_w
)


# Perform backward stepwise regression
ols2_stepwise_w <- step(
  ols2_full_w, 
  direction = "backward",
  trace = 1 # Prints the stepwise regression process
)

# Summary of the final stepwise model
summary(ols2_stepwise_w)

# Add predictions from the final stepwise model
df_merged_w <- add_predictions(ols2_stepwise_w, df_merged_w, "predicted_sales2")

# Plot Actual vs Predicted Values
ggplot(df_merged_w, aes(x = week)) +
  geom_line(aes(y = exp(sales_w), color = "Actual Sales"), size = 1) +
  geom_line(aes(y = exp(predicted_sales0), color = "Model 0"), linetype = "dashed", size = 1) +
  geom_line(aes(y = exp(predicted_sales1), color = "Model 1"), linetype = "dotted", size = 1) +
  geom_line(aes(y = exp(predicted_sales2), color = "Model 2 Stepwise"), linetype = "dotdash", size = 1) +
  labs(title = "Actual vs Predicted Weekly Sales for All Models",
       x = "Week", y = "Sales", color = "Legend") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Models to compare
models_w <- list(
  "Model trend" = ols0w,
  "Model trend + season" = ols1w,
  "Model all covariates step" = ols2_stepwise_w
)

# Get R^2 and AIC for each model
model_stats_w <- get_model_stats(models_w)

# View the results
print(model_stats_w)


rmse_stats_w <- data.frame(
  Model = character(),
  RMSE = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each model
for (i in seq_along(models_w)) {
  model_name <- names(models_w)[i]
  predicted_column <- paste0("predicted_sales", i - 1)  # Adjust column name index
  
  # Calculate RMSE on the original scale
  rmse <- calculate_rmse(
    observed = exp(df_merged_w$sales_w),          # Exponentiate actual values
    predicted = exp(df_merged_w[[predicted_column]])  # Exponentiate predicted values
  )
  
  # Append results to the RMSE stats table
  rmse_stats_w <- rbind(rmse_stats_w, data.frame(Model = model_name, RMSE = rmse))
}

# View RMSE statistics
print(rmse_stats_w)

```

```{r}
rmse_ols_w <- rmse_stats_w$RMSE[3]
rmse_ols_w
```


```{r}
# Daily Models
head(df_merged_d,25)
# properly start in december
df_merged_d <-  df_merged_d %>%
  filter(date > "2021-11-30")
head(df_merged_d)

## Model 0: Trend only
ols0d <- run_model(sales_cop ~ numeric_day, df_merged_d, "Model 0A")
df_merged_d <- add_predictions(ols0d, df_merged_d, "predicted_sales0")

## Model 1: Trend + Seasonality
ols1d <- run_model(sales_cop ~ numeric_day + seasonal_month + day_of_week, df_merged_d, "Model 1A")
df_merged_d <- add_predictions(ols1d, df_merged_d, "predicted_sales1")

# Model 2: Backward
head(df_merged_d)

# Start with the full model (excluding food and bar)
ols2_full_d <- lm(
  sales_cop ~ numeric_day + seasonal_month + day_of_week + fx +
     tmedian + rain_sum, 
  data = df_merged_d
)
summary(ols2_full_d)

# Perform backward stepwise regression
ols2_stepwise_d <- step(
  ols2_full_d, 
  direction = "backward",
  trace = 1 # Prints the stepwise regression process
)

# Summary of the final stepwise model
summary(ols2_stepwise_d)

# Add predictions from the final stepwise model
df_merged_d <- add_predictions(ols2_stepwise_d, df_merged_d, "predicted_sales2")

# Plot Actual vs Predicted Values
ggplot(df_merged_d, aes(x = date)) +
  geom_line(aes(y = exp(sales_cop), color = "Actual Sales"), size = 1) +
  geom_line(aes(y = exp(predicted_sales0), color = "Model 0"), linetype = "dashed", size = 1) +
  geom_line(aes(y = exp(predicted_sales1), color = "Model 1"), linetype = "dotted", size = 1) +
  geom_line(aes(y = exp(predicted_sales2), color = "Model 2 Stepwise"), linetype = "dotdash", size = 1) +
  labs(title = "Actual vs Predicted Sales for All Models",
       x = "date", y = "Sales", color = "Legend") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Models to compare
models_d <- list(
  "Model trend" = ols0d,
  "Model trend + season" = ols1d,
  "Model all covariates step" = ols2_stepwise_d
)

# Get R^2 and AIC for each model
model_stats_d <- get_model_stats(models_d)

# View the results
print(model_stats_d)

#  RMSE calculation for the original (exponentiated) scale for daily models
rmse_stats_d <- data.frame(
  Model = character(),
  RMSE = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each model
for (i in seq_along(models_d)) {
  model_name <- names(models_d)[i]
  predicted_column <- paste0("predicted_sales", i - 1)  # Adjust column name index
  
  # Calculate RMSE on the original scale
  rmse <- calculate_rmse(
    observed = exp(df_merged_d$sales_cop),          # Exponentiate actual values
    predicted = exp(df_merged_d[[predicted_column]])  # Exponentiate predicted values
  )
  
  # Append results to the RMSE stats table
  rmse_stats_d <- rbind(rmse_stats_d, data.frame(Model = model_name, RMSE = rmse))
}

# View RMSE statistics for daily data
print(rmse_stats_d)
```

```{r}
rmse_ols_d <- rmse_stats_d$RMSE[3]
rmse_ols_d
```

## Non Linear Models
Here we explore non linear models, starting from the simplest to more elaborate models in the end combining some of the used models.

### Wrangle data for models
The time series are altered so the visualizations are more understandable, basically we change the date index in the timeseries objects

```{r}
# re-declare time-series beacause we droped some rows:
# Ensure the 'date' columns are in Date format
df_merged_d$date <- as.Date(df_merged_d$date)
df_merged_w$date <- as.Date(df_merged_w$week)
df_merged_m$date <- as.Date(df_merged_m$month)

# Extract the start date and year for each dataframe
start_d <- min(df_merged_d$date)
start_w <- min(df_merged_w$date)
start_m <- min(df_merged_m$date)

# Extract components for daily, weekly, and monthly start times
start_d_year <- as.numeric(format(start_d, "%Y"))
start_d_day <- as.numeric(format(start_d, "%j")) # Day of the year

start_w_year <- as.numeric(format(start_w, "%Y"))
start_w_week <- as.numeric(format(start_w, "%U")) + 1 # Week number, adding 1 since R starts at week 0

start_m_year <- as.numeric(format(start_m, "%Y"))
start_m_month <- as.numeric(format(start_m, "%m"))

# Declare time series with appropriate frequencies
sales_d_ts <- ts(exp(df_merged_d$sales_cop), start = c(start_d_year, start_d_day), frequency = 365)
sales_w_ts <- ts(exp(df_merged_w$sales_w), start = c(start_w_year, start_w_week), frequency = 52)
sales_m_ts <- ts(exp(df_merged_m$sales_m), start = c(start_m_year, start_m_month), frequency = 12)

food_d_ts <- ts(exp(df_merged_d$food), start = c(start_d_year, start_d_day), frequency = 365)
food_w_ts <- ts(exp(df_merged_w$food_w), start = c(start_w_year, start_w_week), frequency = 52)
food_m_ts <- ts(exp(df_merged_m$food_m), start = c(start_m_year, start_m_month), frequency = 12)

bar_d_ts <- ts(exp(df_merged_d$bar), start = c(start_d_year, start_d_day), frequency = 365)
bar_w_ts <- ts(exp(df_merged_w$bar_w), start = c(start_w_year, start_w_week), frequency = 52)
bar_m_ts <- ts(exp(df_merged_m$bar_m), start = c(start_m_year, start_m_month), frequency = 12)

```

```{r}
# Verify the created time series
par(mfrow=c(1,1))
plot(sales_d_ts)
plot(sales_w_ts)
plot(sales_m_ts)


plot(food_d_ts)
plot(food_w_ts)
plot(food_m_ts)

plot(bar_d_ts)
plot(bar_w_ts)
plot(bar_m_ts)
```

Here we fill the sales = 0 values with the mean of the two adjacent dates. This in order to have smoother models.
The dates with sales = 0 are dates that are national holiday like christmas or new years, or inventory day in which the kitchen cannot operate so the sales are 0.
```{r}
# Function to replace 1s with the mean of previous and next observations
fill_ones <- function(ts_data) {
  # Convert time series to numeric vector
  ts_vec <- as.numeric(ts_data)
  
  # Loop through and replace 1s
  for (i in seq_along(ts_vec)) {
    if (ts_vec[i] == 1) {
      # Check boundaries to avoid indexing issues
      prev_val <- ifelse(i > 1, ts_vec[i - 1], NA)
      next_val <- ifelse(i < length(ts_vec), ts_vec[i + 1], NA)
      
      # Replace with mean of previous and next, ignoring NA
      ts_vec[i] <- mean(c(prev_val, next_val), na.rm = TRUE)
    }
  }
  
  # Return as time series with original attributes
  ts(ts_vec, start = start(ts_data), frequency = frequency(ts_data))
}

# Apply the function 
sales_d_ts <- fill_ones(sales_d_ts)
sales_w_ts <- fill_ones(sales_w_ts)
sales_m_ts <- fill_ones(sales_m_ts)


food_d_ts <- fill_ones(food_d_ts)
food_w_ts <- fill_ones(food_w_ts)
food_m_ts <- fill_ones(food_m_ts)

bar_d_ts <- fill_ones(bar_d_ts)
bar_w_ts <- fill_ones(bar_w_ts)
bar_m_ts <- fill_ones(bar_m_ts)

```

### Bass Model

```{r}
# Some simple plots
plot(sales_m_ts)
plot(cumsum(sales_m_ts)) #Returns a vector whose elements are the cumulative sums
```

```{r}
# Bass model
bm_m<-BM(sales_m_ts,display = T) # show graphical view of results / display = True

summary(bm_m)
```

```{r}
bm_m$coefficients['m'] - sum(sales_m_ts)
```
According to this, there are only 1m cop left to sell, this is less than a year / seems wrong. Fits well but the 30- onward is wierd + sales might not be declining yet. Still reflects the innovation and copying in some sense
 
Also the restaurants rely in word of mouth to reach full stage
m = 4.664.000.000 COP, i.e 1 mm EUR approx. / The restaurant has sold 3.515.788.885/ According to this only in 1 year it should extinguish sells p, innovation: 0.832% indicates that the adoption rate due to external influence is relatively low, but not uncommon for many markets. - it is actually relativly innovative q: (8.96%) suggests that imitation plays a larger role than innovation in driving adoption in this market



```{r}
pred_bm_m<- predict(bm_m, newx=c(1:length(sales_m_ts)))
pred_bm_m <- ts(pred_bm_m, start = start(sales_m_ts), frequency = frequency(sales_m_ts))
pred.inst_bm_m <- make.instantaneous(pred_bm_m)
pred.inst_bm_m <- ts(pred.inst_bm_m, start = start(sales_m_ts), frequency = frequency(sales_m_ts))

# plot
plot(sales_m_ts, type = "p", col = "black", pch = 16, cex = 0.7,
     xlab = "Month", ylab = "Monthly Sales", main = "Actual vs Fitted Sales")

# Add the fitted values as a line
lines(pred.inst_bm_m, col = "red", lwd = 2)

# Add a legend
legend("topleft", legend = c("Actual Values", "Fitted Values"),
       col = c("black", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))
```

```{r}
# check residuals
res_bm_m <- sales_m_ts - pred.inst_bm_m
tsdisplay(res_bm_m)
```
Residuals have some structure and 2 lag has correlation.

```{r}
# Calculate RMSE for Bass Model predictions
rmse_bm_m <- calculate_rmse(observed = sales_m_ts, predicted = pred.inst_bm_m)

# Print the RMSE
cat("RMSE for Bass Model Predictions:", rmse_bm_m, "\n")

```

```{r}
bm_w<-BM(sales_w_ts,display = T) # show graphical view of results / display = True
summary(bm_w)
bm_w$coefficients['m'] - sum(sales_w_ts)
# results are similar in terms of m, p and w are in other scale 
#because they are in different time stamp
bm_m$coefficients['q'] / bm_w$coefficients['q'] # they are approx 4 times
bm_m$coefficients['p'] / bm_w$coefficients['p'] # they are approx 4 times
# which makes sense
```
Coefficients are approximatly 4 times the ones of the monthly model, making sense because there are 4 weeks in a month. While market potential is similar.

```{r}
# Prediction
pred_bm_w<- predict(bm_w, newx=c(1:length(sales_w_ts)))
pred_bm_w <- ts(pred_bm_w, start = start(sales_w_ts), frequency = frequency(sales_w_ts))
pred.inst_bm_w <- make.instantaneous(pred_bm_w)
pred.inst_bm_w <- ts(pred.inst_bm_w, start = start(sales_w_ts), frequency = frequency(sales_w_ts))

# plot
plot(sales_w_ts, type = "p", col = "black", pch = 16, cex = 0.7,
     xlab = "Week", ylab = "Weekly Sales", main = "Actual vs Fitted Sales")

# Add the fitted values as a line
lines(pred.inst_bm_w, col = "red", lwd = 2)

# Add a legend
legend("topleft", legend = c("Actual Values", "Fitted Values"),
       col = c("black", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))

```


```{r}
# check residuals
res_bm_w <- sales_w_ts - pred.inst_bm_w
tsdisplay(res_bm_w)
```

Residuals have some structure and 2 lag has correlation, with clear trend and structure in the residuals

```{r}
# RMSE
# Calculate RMSE for Bass Model predictions
rmse_bm_w <- calculate_rmse(observed = sales_w_ts, predicted = pred.inst_bm_w)

# Print the RMSE
cat("RMSE for Bass Model Predictions:", rmse_bm_w, "\n")

```
```{r}
bm_d <- BM(
  sales_d_ts,
  prelimestimates = c(1.2 * sum(sales_d_ts), 0.005, 0.5), # Adjust these estimates
  display = TRUE
)


summary(bm_d)
bm_d$coefficients['m'] - sum(sales_d_ts)
# results are similar in terms of m, p and w are in other scale 
#because they are in different time stamp
bm_w$coefficients['q'] / bm_d$coefficients['q'] # they are approx 7 times
bm_w$coefficients['p'] / bm_d$coefficients['p'] # they are approx 7 times

```
Coefficients are approximately 1:7 scale of the ones in the weekly model, making sense.
The market potential is also similar in order of magnitude.

```{r}
# Prediction
pred_bm_d <- predict(bm_d, newx = c(1:length(sales_d_ts)))
pred_bm_d <- ts(pred_bm_d, start = start(sales_d_ts), frequency = frequency(sales_d_ts))
pred.inst_bm_d <- make.instantaneous(pred_bm_d)
pred.inst_bm_d <- ts(pred.inst_bm_d, start = start(sales_d_ts), frequency = frequency(sales_d_ts))

# Plot actual vs fitted sales for daily data
plot(sales_d_ts, type = "p", col = "black", pch = 16, cex = 0.7,
     xlab = "Day", ylab = "Daily Sales", main = "Actual vs Fitted Sales (Daily)")

# Add the fitted values as a line
lines(pred.inst_bm_d, col = "red", lwd = 2)

# Add a legend
legend("topleft", legend = c("Actual Values", "Fitted Values"),
       col = c("black", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))

```


```{r}
# Check residuals
res_bm_d <- sales_d_ts - pred.inst_bm_d
tsdisplay(res_bm_d)
```

Residuals don not seem stationary, or at least they have a lot of autocorrelation.

```{r}
# Calculate RMSE for Bass Model predictions (daily data)
rmse_bm_d <- calculate_rmse(observed = sales_d_ts, predicted = pred.inst_bm_d)

# Print the RMSE
cat("RMSE for Daily Bass Model Predictions:", rmse_bm_d, "\n")

```



#### Limitation of of the Bass Model

-   Bass model assumes that every product succeeds and the sales
    saturate to the steady state level. However, most new products fail
    in reality.

-   The market potential *m* is constant along the whole life cycle.

-   Bass model predictions works well only after the scale inflection
    point. if sales of a category goes up and up like a J-curve, it can
    over estimate the overall market size.

-   It is a model for products with a limited life cycle: needs a
    hypothesis.

-   Another drawback of Bass model is that the diffusion pattern in not
    affected by marketing mix variables like price or advertising.

The generalized Bass model extends the original Bass model allowing the
roles of marketing mix value.

### Generalized Bass Model

Bass model is used to forecast the adoption of a new product and to
predict the sales, since it determines the shape of the curve of a model
that represent the cumulative adoption of a new product. The Generalized
Bass model extends the original Bass model by incorporating marketing
mix variables. We can know the effect of pricing, promotions on the new
product diffusion curve. It is more flexible than the original Bass
model.

```{r}

m <- 4.451570e+09
p <- 8.472917e-03
q <- 9.415625e-02

GBM_monthly_sales <- GBM(
  sales_m_ts, 
  shock = 'exp', 
  nshock = 1,
  #prelimestimates = c(m,p,q, 12, 0.1, -0.1)
  prelimestimates = c(m,p,q, 10, 0.1, 2)
  #prelimestimates = c(m,p,q, 11, 15, -0.1)
  )

summary(GBM_monthly_sales)

pred_GBM_monthly_sales<- predict(GBM_monthly_sales, newx=c(1:60))
pred_GBM_monthly_sales.inst<- make.instantaneous(pred_GBM_monthly_sales)
```

### Guseo-Guidolin Model

```{r}
# Montly model
ggm1 <- GGM(sales_m_ts, mt='base', display = T)
ggm2 <- GGM(sales_m_ts, mt= function(x) pchisq(x,10),display = T)
summary(ggm1)
summary(ggm2)
# try different functions for market potential

ggm3 <- GGM(sales_m_ts, mt= function(x) log(x),display = T)
ggm4 <- GGM(sales_m_ts, mt= function(x) (x)**(1/1.05),display = T)
summary(ggm3)
summary(ggm4)
```



K \<- 7.683785e+09

pc \<- 2.698613e-02

qc \<- 2.582412e-01

ps \<- 7.731763e-03

qs \<- 4.508202e-02

```{r}
# predictions
pred_ggm_m <- predict(ggm1, newx = c(1:length(sales_m_ts)))
pred_ggm_m <- ts(pred_ggm_m, start = start(sales_m_ts), frequency = frequency(sales_m_ts))
pred.inst_ggm_m <- make.instantaneous(pred_ggm_m)
pred.inst_ggm_m <- ts(pred.inst_ggm_m, start = start(sales_m_ts), frequency = frequency(sales_m_ts))

# Plot actual vs fitted sales for monthly data
plot(sales_m_ts, type = "p", col = "black", pch = 16, cex = 0.7,
     xlab = "Month", ylab = "Monthly Sales", main = "Actual vs Fitted Sales (GGM Model)")

# Add the fitted values as a line
lines(pred.inst_ggm_m, col = "red", lwd = 2)

# Add a legend
legend("topleft", legend = c("Actual Values", "Fitted Values (GGM)"),
       col = c("black", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))

```


```{r}
#Analysis of residuals
res_GGM_m<- sales_m_ts - pred.inst_ggm_m
tsdisplay(res_GGM_m)
```
Residuals look stationary for this model

```{r}
# Residuals somehow are kind of stationary
# check for stationarity of residuals
adf_test <- adf.test(res_GGM_m)
print(adf_test) # if p-val < alpha, series stationary
# so with this model we achieve stationary series

# check for autocorrelation in residuals
Box.test(res_GGM_m, lag = 10, type = "Ljung-Box") # h0 res indep
# p-val > alpha => fail to reject h0, so residuals seem indep

```
Residuals are likeley stationary

```{r}
# Calculate RMSE for ggm1
rmse_ggm1 <- calculate_rmse(observed = sales_m_ts, predicted = pred.inst_ggm_m)

# Print RMSE for ggm1
cat("RMSE for GGM Model 1 (Base):", rmse_ggm1, "\n")

```
```{r}
# Weekly
ggm1_w <- GGM(sales_w_ts, mt='base', display = T)
ggm2_w <- GGM(sales_w_ts, mt= function(x) pchisq(x,25),display = T)
summary(ggm1_w) # this one is better
summary(ggm2_w)
# try different functions for market potential

ggm3_w <- GGM(sales_w_ts, mt= function(x) log(x),display = T)
ggm4_w <- GGM(sales_w_ts, mt= function(x) (x)**(1/1.05),display = T)

summary(ggm3_w)
summary(ggm4_w) # better shaped but less significant

```

```{r}
# predictions
pred_ggm_w <- predict(ggm1_w, newx = c(1:length(sales_w_ts)))
pred_ggm_w <- ts(pred_ggm_w, start = start(sales_w_ts), frequency = frequency(sales_w_ts))
pred.inst_ggm_w <- make.instantaneous(pred_ggm_w)
pred.inst_ggm_w <- ts(pred.inst_ggm_w, start = start(sales_w_ts), frequency = frequency(sales_w_ts))

# Plot actual vs fitted sales for weekly data
plot(sales_w_ts, type = "p", col = "black", pch = 16, cex = 0.7,
     xlab = "Week", ylab = "Weekly Sales", main = "Actual vs Fitted Sales (GGM Model)")

# Add the fitted values as a line
lines(pred.inst_ggm_w, col = "red", lwd = 2)

# Add a legend
legend("topleft", legend = c("Actual Values", "Fitted Values (GGM)"),
       col = c("black", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))

```

```{r}
# Analysis of residuals
res_GGM_w <- sales_w_ts - pred.inst_ggm_w
tsdisplay(res_GGM_w)


# Check for stationarity of residuals
adf_test_w <- adf.test(res_GGM_w)
print(adf_test_w) # if p-value < alpha, series is stationary

# Check for autocorrelation in residuals
box_test_w <- Box.test(res_GGM_w, lag = 10, type = "Ljung-Box")
print(box_test_w) # if p-value > alpha, residuals are independent

```
Series is stationary according to tests, but clearly has strong autocorrelation

```{r}
# RMSE
rmse_ggm_w <- calculate_rmse(observed = sales_w_ts, predicted = pred.inst_ggm_w)

# Print the RMSE
cat("RMSE for Weekly GGM Model Predictions:", rmse_ggm_w, "\n")

```

```{r}
# Daily GGM
# Scaling the sales data
sales_min <- min(sales_d_ts)
sales_max <- max(sales_d_ts)
sales_scaled <- (sales_d_ts - sales_min) / (sales_max - sales_min)

# View scaled data
summary(sales_scaled)
plot(sales_scaled, type = "l", main = "Scaled Daily Sales", xlab = "Day", ylab = "Scaled Sales")

```
We re-scale the data because else the model won't converge

```{r}
# Fit GGM models using scaled data
ggm1_d <- GGM(sales_scaled, mt = 'base', display = T)
ggm2_d <- GGM(sales_scaled, mt = function(x) pchisq(x, 10), display = T)
ggm3_d <- GGM(sales_scaled, mt = function(x) log(x), display = T)
ggm4_d <- GGM(sales_scaled, mt = function(x) (x)^(1/1.05), display = T)

# Summarize models
summary(ggm1_d)  # Base model
summary(ggm2_d)  # Chi-squared
summary(ggm3_d)  # Log transformation
summary(ggm4_d)  # Power transformation

```

Predict on the best model based on fit and p-values
We select model 1
```{r}
# Prediction using GGM model
pred_ggm_d <- predict(ggm1_d, newx = c(1:length(sales_scaled)))
pred_ggm_d <- ts(pred_ggm_d, start = start(sales_scaled), frequency = frequency(sales_scaled))
pred.inst_ggm_d <- make.instantaneous(pred_ggm_d)
pred.inst_ggm_d <- ts(pred.inst_ggm_d, start = start(sales_scaled), frequency = frequency(sales_scaled))

# Re-scale predictions back to the original scale
pred_original_scale <- (pred.inst_ggm_d * (sales_max - sales_min)) + sales_min

# Plot actual vs fitted sales (original scale)
plot(sales_d_ts, type = "p", col = "black", pch = 16, cex = 0.7,
     xlab = "Day", ylab = "Daily Sales", main = "Actual vs Fitted Sales (Original Scale)")
lines(pred_original_scale, col = "red", lwd = 2)
legend("topleft", legend = c("Actual Values", "Fitted Values (GGM, Original Scale)"),
       col = c("black", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))

```

```{r}
# Analysis of residuals
res_GGM_d <- sales_d_ts - pred_original_scale
tsdisplay(res_GGM_d, main = "Residuals of GGM Model")

```
Residuals dont look stationary

```{r}
# Check for stationarity of residuals
adf_test_d <- adf.test(res_GGM_d)
print(adf_test_d)  # If p-value < alpha, series is stationary
# according to this, they are stationary

# Check for autocorrelation in residuals
box_test_d <- Box.test(res_GGM_d, lag = 10, type = "Ljung-Box")
print(box_test_d)  # If p-value > alpha, residuals are indep

```
Residuals look stationary in the test but hey have serial correlation

```{r}
# Calculate RMSE for GGM model predictions (original scale)
rmse_ggm_d <- calculate_rmse(observed = sales_d_ts, predicted = pred_original_scale)

# Print the RMSE
cat("RMSE for Daily GGM Model Predictions (Original Scale):", rmse_ggm_d, "\n")

```

### Holt-Winters Model

```{r}
# adjust timeseries to ensure date consistency
sales_m_ts <- ts(sales_m_ts, frequency=12, start=c(2021, 11))

hw1_m<- hw(sales_m_ts, seasonal="additive")
hw2_m<- hw(sales_m_ts, seasonal="multiplicative")

# prediction
fitted_hw1 <- hw1_m$fitted
fitted_hw2 <- hw2_m$fitted

```

We now plot the models

```{r}
# Create a data frame for ggplot
plot_data <- data.frame(
  Time = time(sales_m_ts),
  Actual = as.numeric(sales_m_ts),
  Fitted_Additive = as.numeric(hw1_m$fitted),
  Fitted_Multiplicative = as.numeric(hw2_m$fitted)
)

# Melt data for easier ggplot usage
library(reshape2)
plot_data_melted <- melt(plot_data, id.vars = "Time", 
                         variable.name = "Series", 
                         value.name = "Value")

# Plot using ggplot2
ggplot(plot_data_melted, aes(x = Time, y = Value, color = Series)) +
  geom_point(data = subset(plot_data_melted, Series == "Actual"), size = 2) + # Actual values as dots
  geom_line(data = subset(plot_data_melted, Series != "Actual"), size = 1) +  # Fitted values as lines
  labs(
    title = "Actual vs Fitted Values",
    x = "Time",
    y = "Value",
    color = "Series"
  ) +
  scale_color_manual(
    values = c("Actual" = "black", "Fitted_Additive" = "blue", "Fitted_Multiplicative" = "red"),
    labels = c("Actual", "Fitted (Additive)", "Fitted (Multiplicative)")
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

```
Looks like the multiplicative models follows the data more more closely in general.

```{r}
# residuals
residuals_hw1 <- residuals(hw1_m)  
residuals_hw2 <- residuals(hw2_m)  
tsdisplay(residuals_hw1)
tsdisplay(residuals_hw2)

```

```{r}
# Stationarity and Correlation
# check for stationarity of residuals
# additive
adf_test <- adf.test(residuals_hw1) # H0: series is non-stationary
print(adf_test) # if p-val < alpha, series not stationary
# so with this model we achieve stationary series
# multiplicative
adf_test <- adf.test(residuals_hw2) # H0: series is non-stationary
print(adf_test) # if p-val < alpha, series not stationary
# so with this model we achieve stationary series

# additive
# check for autocorrelation in residuals
Box.test(residuals_hw1, lag = 10, type = "Ljung-Box") # h0 res indep
# p-val > alpha =>  Dont reject h0, so residuals are indep

# additive
# check for autocorrelation in residuals
Box.test(residuals_hw2, lag = 10, type = "Ljung-Box") # h0 res indep
# p-val > alpha =>  Dont reject h0, so residuals are indep

```
Multiplicative model follows the data better and has slightly better residuals

```{r}
# forecast
# save the forecast of the second model
forecast_hw1 <- forecast(hw1_m, h=12)
forecast_hw2 <- forecast(hw2_m, h=12)

# Forecast plot
# Plot the time series with both forecasts
autoplot(sales_m_ts) +
  autolayer(forecast_hw1$mean, series="Additive Holt-Winters Forecast", PI=F) +
  autolayer(forecast_hw2$mean, series="Multiplicative Holt-Winters Forecast", PI=F) +
  ggtitle("Sales Forecast with Holt-Winters Models") +
  xlab("Time") +
  ylab("Sales") +
  scale_color_manual(
    values=c("Additive Holt-Winters Forecast" = "blue",
             "Multiplicative Holt-Winters Forecast" = "red")
  ) +
  theme_minimal() +
  theme(legend.position = "top", legend.title = element_blank())

```

```{r}
# RMSE Calculation for Holt-Winters models
rmse_hw1 <- calculate_rmse(observed = sales_m_ts, predicted = fitted_hw1)
rmse_hw2 <- calculate_rmse(observed = sales_m_ts, predicted = fitted_hw2)

# Print RMSE values
cat("RMSE for Additive Holt-Winters Model:", rmse_hw1, "\n")
cat("RMSE for Multiplicative Holt-Winters Model:", rmse_hw2, "\n")

```
Multiplicative model is better

The Holt winters model has a frequency limit of 24, so we cannot do larger than that. Weekly and daily data have 52 and 365 frequencies respectively so we cannot fit the model with the R implementation so far.



### ARIMA models

ARIMA is a acronym for Auto Regressive Integrated Moving Average, ARIMA
(p,d,q) where p refers to the AR part, q refers to the MA part and d is
the degree of first difference involved.


First we nwwd to check if the series is stationary
```{r}
# see if series is stationary
adf.test(sales_m_ts) #H0, series is non-stationary
# p-val > 0.05 => dont reject, non stationary: series is not stationary
adf.test(diff(sales_m_ts)) #H0, series is non-stationary

# see the acf and pacf
tsdisplay(diff(sales_m_ts))

```

#### Monthly sales

```{r}
plot(sales_m_ts)
```

```{r}
ndiffs(sales_m_ts)

```

```{r}
tsdisplay(diff(sales_m_ts))
```

Correlogram plot maybe suggests AR-1 or MA-1 after first difference

```{r}
# ARIMA(p,d,q) = (1,1,0)
arima1_m<- Arima(sales_m_ts, order=c(1,1,0))
summary(arima1_m)

```
```{r}
# study residual to see if is a good model
resid1_m<- residuals(arima1_m)
tsdisplay(resid1_m)
```

Residuals look stationary after fitting ARIMA

```{r}
auto_arima_m <- auto.arima(sales_m_ts)
auto_arima_m

autoplot(forecast(auto_arima_m))
checkresiduals(auto_arima_m)
```

The residuals of the Autoarima look stationary

AIC of the the manual arima is 1265, while the one of the autoarima is 1263. Lets use the autoarima

```{r}
# Fitted values from both models
fitted_auto_arima <- fitted(auto_arima_m)
fitted_arima1 <- fitted(arima1_m)

# Create a data frame for plotting
plot_data <- data.frame(
  Time = time(sales_m_ts),
  Actual = as.numeric(sales_m_ts),
  Fitted_Auto_ARIMA = as.numeric(fitted_auto_arima),
  Fitted_ARIMA1 = as.numeric(fitted_arima1)
)

# Melt the data frame
plot_data_melted <- melt(plot_data, id.vars = "Time", 
                         variable.name = "Series", 
                         value.name = "Value")

# Plot

ggplot(plot_data_melted, aes(x = Time, y = Value, color = Series)) +
  geom_point(data = subset(plot_data_melted, Series == "Actual"), size = 2) +  # Actual values as points
  geom_line(data = subset(plot_data_melted, Series != "Actual"), size = 1) +   # Fitted values as lines
  labs(
    title = "Actual vs Fitted Values for ARIMA Models",
    x = "Time",
    y = "Sales",
    color = "Series"
  ) +
  scale_color_manual(
    values = c("Actual" = "black", "Fitted_Auto_ARIMA" = "blue", "Fitted_ARIMA1" = "red"),
    labels = c("Actual", "Fitted (Auto ARIMA)", "Fitted (ARIMA(1,1,0))")
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )


```


```{r}
# Calculate RMSE for the fitted values
# Calculate RMSE for each model
rmse_auto_arima <- calculate_rmse(observed = sales_m_ts, predicted = fitted_auto_arima)
rmse_arima1 <- calculate_rmse(observed = sales_m_ts, predicted = fitted_arima1)

# Print RMSE values
cat("RMSE for Auto ARIMA Model:", rmse_auto_arima, "\n")
cat("RMSE for ARIMA(1,1,0) Model:", rmse_arima1, "\n")


```
The RMSE of the Autoarima is better as is the AIC.

The ARIMA(0,1,1) model can be described simply as a **random walk with
drift**. Here's what that means:

1.  **AR (AutoRegressive) Part:**

    -   The first number, **0**, indicates the order of the
        autoregressive part. In this case, it means there are no
        autoregressive terms (i.e., the model does not use past values
        of the series to predict future values).

2.  **I (Integrated) Part:**

    -   The second number, **1**, indicates the degree of differencing
        required to make the time series stationary. Differencing is a
        technique used to remove trends and seasonality from the series.
        A value of 1 means the series is differenced once.

3.  **MA (Moving Average) Part:**

    -   The third number, **1**, indicates the order of the moving
        average part. 

An ARIMA(0,1,1) model is suitable when:

The d=1 parameter in ARIMA(0,1,1) indicates that the series is differenced once to achieve stationarity.
Before differencing, the series may exhibit a linear trend or random walk behavior.
After differencing, the series should show no trend and have relatively stable mean and variance

The q=1 in ARIMA(0,1,1) indicates that the series is modeled with a first-order moving average component after differencing.
The autocorrelation function (ACF) of the differenced series should show:
A significant spike at lag 1.
Rapid decay to zero after lag 1.
The partial autocorrelation function (PACF) should show no significant lags.


```{r}
# study residual to see if is a good model
resid_autoarima_m<- residuals(auto_arima_m)
tsdisplay(resid_autoarima_m)
```

#### Weekly sales



```{r}
# see if series is stationary
adf.test(sales_w_ts) #H0, series is non-stationary
# p-val > 0.05 => dont reject, non stationary: series is not stationary
adf.test(diff(sales_w_ts)) # after diff is sationary
```
After differencing, looks stationary

```{r}
tsdisplay(diff(sales_w_ts))
```

Correlograms suggest maybe AR 1 or MA 1.

```{r}
### Manual ARIMA------------
# ARIMA(p,d,q) = (1,1,0)
arima1_w<- Arima(sales_w_ts, order=c(1,1,0))
summary(arima1_w)

```

```{r}
auto_arima_w <- auto.arima(sales_w_ts)
summary(auto_arima_w)
```

AIC on the Autoarima is better, lets go with that one

```{r}
checkresiduals(auto_arima_w)
```
Residuals look stationary, see the plots for both models

```{r}
# Fit ARIMA models for weekly data
arima1_w <- Arima(sales_w_ts, order = c(1, 1, 0))
auto_arima_w <- auto.arima(sales_w_ts)

# Extract fitted values for both models
fitted_arima1_w <- fitted(arima1_w)
fitted_auto_arima_w <- fitted(auto_arima_w)

# Create a data frame for plotting
plot_data <- data.frame(
  Time = time(sales_w_ts),
  Actual = as.numeric(sales_w_ts),
  Fitted_ARIMA1 = as.numeric(fitted_arima1_w),
  Fitted_Auto_ARIMA = as.numeric(fitted_auto_arima_w)
)

# Melt the data frame for ggplot2
plot_data_melted <- melt(plot_data, id.vars = "Time", 
                         variable.name = "Series", 
                         value.name = "Value")

# Plot using ggplot2
ggplot(plot_data_melted, aes(x = Time, y = Value, color = Series)) +
  geom_point(data = subset(plot_data_melted, Series == "Actual"), size = 2) +  # Actual values as points
  geom_line(data = subset(plot_data_melted, Series != "Actual"), size = 1) +   # Fitted values as lines
  labs(
    title = "Actual vs Fitted Values for ARIMA Models (Weekly)",
    x = "Time",
    y = "Sales",
    color = "Series"
  ) +
  scale_color_manual(
    values = c("Actual" = "black", "Fitted_ARIMA1" = "red", "Fitted_Auto_ARIMA" = "blue"),
    labels = c("Actual", "Fitted (ARIMA(1,1,0))", "Fitted (Auto ARIMA)")
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

```{r}
# Calculate RMSE for both models
rmse_arima1_w <- calculate_rmse(observed = sales_w_ts, predicted = fitted_arima1_w)
rmse_auto_arima_w <- calculate_rmse(observed = sales_w_ts, predicted = fitted_auto_arima_w)

# Print RMSE values
cat("RMSE for ARIMA(1,1,0) Model (Weekly):", rmse_arima1_w, "\n")
cat("RMSE for Auto ARIMA Model (Weekly):", rmse_auto_arima_w, "\n")
```
The Auto-arima is also better in terms of RMSE


#### Daily sales


```{r}
# see if series is stationary
adf.test(sales_d_ts) #H0, series is non-stationary
# p-val < 0.05 =>  reject non stationary: series might be stationary

```
No need for differencing because is already stationary, try to model with arima

```{r}
tsdisplay(sales_d_ts)
```


Autocorrelograms are not easy to interpret, but lets try with a baseline model

```{r}
# ARIMA(p,d,q) = (2,1,0)
arima1_d<- Arima(sales_d_ts, order=c(1,0,1))
summary(arima1_d)

```
```{r}
checkresiduals(arima1_d)
```

Residuals look not entirely stationary

Try to model with automatic approach:

```{r}
auto_arima_d <- auto.arima(sales_d_ts)
summary(auto_arima_d)
```


```{r}
checkresiduals(auto_arima_d)
```

Rresiduals improve, and AIC is lower in the autoarima
Check the fit for both models

```{r}
# Extract fitted values for both models
fitted_arima1_d <- fitted(arima1_d)
fitted_auto_arima_d <- fitted(auto_arima_d)

# Create a data frame for plotting
plot_data <- data.frame(
  Time = time(sales_d_ts),
  Actual = as.numeric(sales_d_ts),
  Fitted_ARIMA1 = as.numeric(fitted_arima1_d),
  Fitted_Auto_ARIMA = as.numeric(fitted_auto_arima_d)
)

# Melt the data frame for ggplot2
plot_data_melted <- melt(plot_data, id.vars = "Time", 
                         variable.name = "Series", 
                         value.name = "Value")

# Plot using ggplot2
ggplot(plot_data_melted, aes(x = Time, y = Value, color = Series)) +
  geom_point(data = subset(plot_data_melted, Series == "Actual"), size = 2) +  # Actual values as points
  geom_line(data = subset(plot_data_melted, Series != "Actual"), size = 1) +   # Fitted values as lines
  labs(
    title = "Actual vs Fitted Values for ARIMA Models (Daily)",
    x = "Time",
    y = "Sales",
    color = "Series"
  ) +
  scale_color_manual(
    values = c("Actual" = "black", "Fitted_ARIMA1" = "red", "Fitted_Auto_ARIMA" = "blue"),
    labels = c("Actual", "Fitted (ARIMA(1,0,1))", "Fitted (Auto ARIMA)")
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

Plot is not readable, but check the RMSE for both models to confirm wihch fits better

```{r}
# Calculate RMSE for both models
rmse_arima1_d <- calculate_rmse(observed = sales_d_ts, predicted = fitted_arima1_d)
rmse_auto_arima_d <- calculate_rmse(observed = sales_d_ts, predicted = fitted_auto_arima_d)

# Print RMSE values
cat("RMSE for ARIMA(1,0,1) Model (Daily):", rmse_arima1_d, "\n")
cat("RMSE for Auto ARIMA Model (Daily):", rmse_auto_arima_d, "\n")
```
Autoarima is much better, now try to improve with seasonality, beacuse daily data looks seasonal each 7 days.

### SARIMA
```{r}
# Daily sales
tsdisplay(sales_d_ts) # 
tsdisplay(diff(sales_d_ts))
```

```{r}
sarima_d <- auto.arima(sales_d_ts, seasonal=TRUE)
summary(sarima_d)
```
```{r}
resid_ds<- residuals(sarima_d)
tsdisplay(resid_ds)

# check for autocorrelation
Box.test(residuals(sarima_d), type="Ljung-Box")
# A low p-value (<0.05) suggests residual autocorrelation.
```


Looks like Aarima is the same in terms of AIC, lets check the RMSE:

```{r}
# Extract fitted values for both models

fitted_sarima_d <- fitted(sarima_d)

# Calculate RMSE for both models
rmse_sarima_d <- calculate_rmse(observed = sales_d_ts, predicted = fitted_sarima_d)

# Print RMSE values
cat("RMSE for Auto ARIMA Model (Daily):", rmse_auto_arima_d, "\n")
cat("RMSE for Seasonal ARIMA Model (Daily):", rmse_sarima_d, "\n")

```
The RMSE is exactly the same, they are the same model.

### SARIMAX
Refine SARIMA with external regressors
```{r}
# readefine sales_d_ts
head(df_merged_d)
sales_d_ts <- ts(exp(df_merged_d$sales_cop), frequency=365, start=c(2021, 334))  # 334 is November 30
seasonal_sales_d_ts <- ts(exp(df_merged_d$sales_cop), frequency=7, start=c(2021, 334))  # 334 is November 30
plot(sales_d_ts)
tsdisplay(sales_d_ts,lag.max = 30)
tsdisplay(seasonal_sales_d_ts,lag.max = 30)
# define regresors
# Select specific columns by name
x_regressors_d <- df_merged_d %>% select(rain_sum, fx, tmedian)
# Apply the exponential function to each column
x_regressors_d <- as.data.frame(apply(x_regressors_d, 2, exp))
# Convert to a matrix for ARIMA modeling
x_regressors_d <- as.matrix(x_regressors_d)

```



```{r}
# fit the model on sales
# Fit an auto.arima model with seasonal component and external regressors
sarimax_model_d <- auto.arima(
  sales_d_ts,
  seasonal = TRUE,               # Enable seasonal components
  xreg = x_regressors_d          # External regressors
)

# Display the summary of the fitted model
summary(sarimax_model_d)

```
The AIC actually decreases, lets check the RMSE

```{r}
# Extract fitted values for all models
fitted_sarimax_d <- fitted(sarimax_model_d)

# Calculate RMSE for all models
rmse_sarimax_d <- calculate_rmse(observed = sales_d_ts, predicted = fitted_sarimax_d)

# Print RMSE values
cat("RMSE for Auto ARIMA Model (Daily):", rmse_auto_arima_d, "\n")
cat("RMSE for Seasonal ARIMA Model (Daily):", rmse_sarima_d, "\n")
cat("RMSE for SARIMAX Model (Daily):", rmse_sarimax_d, "\n")

```
The RMSE also worsens, so stay with regular Auto-ARIMA

### Exponential Smoothing methods

#### Simple Exponential Smoothing

##### Monthly Sales

Monthly Sales

```{r}
fit_m1 <- ses(sales_m_ts, alpha = 0.2, initial = 'simple', h=5)
fit_m2 <- ses(sales_m_ts, alpha = 0.6, initial = 'simple', h=5)
fit_m3 <- ses(sales_m_ts, h=5)

plot(sales_m_ts, ylab='Monthly Sales', xlab='Months')
lines(fitted(fit_m1), col='blue', type='o')
lines(fitted(fit_m2), col='red', type='o')
lines(fitted(fit_m3), col='green', type='o')
```

```{r}
forecast_m1 <- ses(sales_m_ts, h=5)

# Accuracy of one-step-ahead training errors
round(accuracy(forecast_m1),2)

summary(forecast_m1)

autoplot(forecast_m1) + autolayer(fitted(forecast_m1),series='Fitted') + ylab("Monthly Sales")+xlab("Months")
```

```{r}
# Extract fitted values for each model
fitted_m1 <- fitted(fit_m1)
fitted_m2 <- fitted(fit_m2)
fitted_m3 <- fitted(fit_m3)

# Calculate RMSE for each model
rmse_m1 <- calculate_rmse(observed = sales_m_ts, predicted = fitted_m1)
rmse_m2 <- calculate_rmse(observed = sales_m_ts, predicted = fitted_m2)
rmse_m3 <- calculate_rmse(observed = sales_m_ts, predicted = fitted_m3)

# Print RMSE values
cat("RMSE for SES Model 1 (alpha = 0.2):", rmse_m1, "\n")
cat("RMSE for SES Model 2 (alpha = 0.6):", rmse_m2, "\n")
cat("RMSE for SES Model 3 (Optimized alpha):", rmse_m3, "\n")

```
```{r}
rmse_exp_sm_m <- rmse_m2
```

##### Weekly Sales

Weekly Sales

For weekly data, exponential smoothing can capture longer-term trends
and seasonal patterns that repeat on a weekly basis. Weekly data can
also have seasonal components related to months, quarters, or years.



```{r}
fit_w1 <- ses(sales_w_ts, alpha = 0.2, initial = 'simple', h=5)
fit_w2 <- ses(sales_w_ts, alpha = 0.6, initial = 'simple', h=5)
fit_w3 <- ses(sales_w_ts, h=5)

plot(sales_w_ts, ylab='Weekly Sales', xlab='Weeks')
lines(fitted(fit_w1), col='blue', type='o')
lines(fitted(fit_w2), col='red', type='o')
lines(fitted(fit_w3), col='green', type='o')
```

```{r}
forecast_w1 <- ses(sales_w_ts, h=5)
round(accuracy(forecast_w1),2)

summary(forecast_w1)

autoplot(forecast_w1) + autolayer(fitted(forecast_w1),series='Fitted') + ylab("Weekly Sales")+xlab("Weeks")
```

```{r}
# Extract fitted values for each model
fitted_w1 <- fitted(fit_w1)
fitted_w2 <- fitted(fit_w2)
fitted_w3 <- fitted(fit_w3)

# Calculate RMSE for each model
rmse_w1 <- calculate_rmse(observed = sales_w_ts, predicted = fitted_w1)
rmse_w2 <- calculate_rmse(observed = sales_w_ts, predicted = fitted_w2)
rmse_w3 <- calculate_rmse(observed = sales_w_ts, predicted = fitted_w3)

# Print RMSE values
cat("RMSE for SES Model 1 (alpha = 0.2):", rmse_w1, "\n")
cat("RMSE for SES Model 2 (alpha = 0.6):", rmse_w2, "\n")
cat("RMSE for SES Model 3 (Optimized alpha):", rmse_w3, "\n")

```
```{r}
rmse_exp_sm_w <- rmse_w3
```

##### Daily sales

Daily Sales

For daily data, exponential smoothing can be used to forecast short-term
trends and seasonal patterns. When applying exponential smoothing to
daily data, you need to consider:

-   **Seasonality**: Daily data often exhibit seasonal patterns, such as
    weekly cycles (e.g., higher sales on weekends).

-   **Holidays and special events**: These can cause irregular patterns
    in daily data that may need to be accounted for.


```{r}

fit_d1 <- ses(sales_d_ts, alpha = 0.2, initial = 'simple', h=5)
fit_d2 <- ses(sales_d_ts, alpha = 0.6, initial = 'simple', h=5)
fit_d3 <- ses(sales_d_ts, h=5)

plot(sales_d_ts, ylab='Daily Sales', xlab='Days')
lines(fitted(fit_d1), col='blue', type='o')
lines(fitted(fit_d2), col='red', type='o')
lines(fitted(fit_d3), col='green', type='o')
```

```{r}
forecast_d1 <- ses(sales_d_ts, h=5)
round(accuracy(forecast_d1),2)

summary(forecast_d1)

autoplot(forecast_d1) + autolayer(fitted(forecast_d1),series='Fitted') + ylab("Daily Sales")+xlab("Days")
```

```{r}
# Extract fitted values for each model
fitted_d1 <- fitted(fit_d1)
fitted_d2 <- fitted(fit_d2)
fitted_d3 <- fitted(fit_d3)

# Calculate RMSE for each model
rmse_d1 <- calculate_rmse(observed = sales_d_ts, predicted = fitted_d1)
rmse_d2 <- calculate_rmse(observed = sales_d_ts, predicted = fitted_d2)
rmse_d3 <- calculate_rmse(observed = sales_d_ts, predicted = fitted_d3)

# Print RMSE values
cat("RMSE for SES Model 1 (alpha = 0.2):", rmse_d1, "\n")
cat("RMSE for SES Model 2 (alpha = 0.6):", rmse_d2, "\n")
cat("RMSE for SES Model 3 (Optimized alpha):", rmse_d3, "\n")

```
```{r}
rmse_exp_sm_d <- rmse_d3
```

### GGM + SARMAX


#### Monthly
```{r}
# GGM part
# Summary of the GGM model
summary(ggm1)  # Assume ggm1_m is the monthly GGM model

# Predictions using GGM
pred_GGM_m <- predict(ggm1, newx = matrix(1:length(sales_m_ts), ncol = 1))
pred_GGM_m.inst <- make.instantaneous(pred_GGM_m)
```
```{r}
# Convert predictions to a time series
start_time_m <- start(sales_m_ts)  # Start time from sales_m_ts
frequency_m <- frequency(sales_m_ts)  # Frequency from sales_m_ts

pred_GGM_m_vec <- unlist(pred_GGM_m.inst)  # Convert predictions to a numeric vector
pred_GGM_m_ts <- ts(pred_GGM_m_vec, start = start_time_m, frequency = frequency_m)
```


```{r}
# Plot actual vs GGM predictions
plot(sales_m_ts, type = "b", xlab = "Month", ylab = "Monthly Sales", pch = 16, lty = 3, cex = 0.6)
lines(pred_GGM_m_ts, col = "red", lty = 2)
```

```{r}
#### SARIMAX Refinement------------------------

# Get fitted values from the GGM model
fit.sales_m <- fitted(ggm1)

# Check length consistency
if (length(fit.sales_m) != length(sales_m_ts)) {
  stop("fit.sales_m and sales_m_ts lengths do not match")
}

```

```{r}
# Scale GGM fitted values and the cumulative sales
fit.sales_m <- scale(fit.sales_m)
sales_m_ts_scaled <- scale(cumsum(sales_m_ts))  # Scale the cumulative sales for convergence
```


```{r}
# Fit SARIMAX with GGM fitted values as regressors
sarima_m <- Arima(
  sales_m_ts_scaled,
  order = c(1, 0, 1),
  seasonal = list(order = c(0, 0, 1), period = 12),  # Monthly seasonality
  xreg = fit.sales_m
)

summary(sarima_m)
```

```{r}
# Reverse scaling for fitted cumulative values
fitted_cumulative <- fitted(sarima_m)

scaling_center <- attr(sales_m_ts_scaled, "scaled:center")
scaling_scale <- attr(sales_m_ts_scaled, "scaled:scale")

fitted_cumulative_original <- fitted_cumulative * scaling_scale + scaling_center

# Convert cumulative fitted values to instantaneous values
fitted_instantaneous <- diff(c(fitted_cumulative_original, NA))  # Add NA to align lengths

# Create a time series object for the fitted instantaneous values
fitted_instantaneous_ts <- ts(
  fitted_instantaneous,
  start = start(sales_m_ts),
  frequency = frequency(sales_m_ts)
)
```

```{r}
# Plot actual vs fitted instantaneous values
plot(sales_m_ts, type = "p", col = "blue", pch = 16,
     main = "Original vs Fitted Instantaneous Values (Monthly)",
     xlab = "Time", ylab = "Instantaneous Values")
lines(fitted_instantaneous_ts, col = "red", lwd = 3, lty = 1)

# Add legend
legend("bottomright", legend = c("Original sales", "Fitted sales"),
       col = c("blue", "red"), lty = c(NA, 1), pch = c(16, NA), lwd = c(NA, 3))
```


```{r}
# Calculate RMSE for fitted_instantaneous_ts against sales_m_ts
rmse_mixture_m <- calculate_rmse(observed = sales_m_ts, predicted = fitted_instantaneous_ts)

# Print the RMSE value
cat("RMSE for Fitted Instantaneous Values (GGM + SARIMAX):", rmse_mixture_m, "\n")
```
```{r}
resid_mixture_m <- sales_m_ts - fitted_instantaneous_ts
tsdisplay(resid_mixture_m)
```
Residuals have autocorrelation at lag 1

#### Weekly

```{r}
#### GGM-------------------------------

summary(ggm1_w) # this one is best model found


pred_GGM_w<- predict(ggm1_w, newx=matrix(1:length(sales_w_ts), ncol=1))
pred_GGM_w.inst<- make.instantaneous(pred_GGM_w)
# set same timeframe for GGM preds
start_time_w <- start(sales_w_ts)  # Get start time from sales_w_ts
frequency_w <- frequency(sales_w_ts)  # Get frequency from sales_w_ts

# Convert pred_GGM to a numeric vector
pred_GGM_w_vec <- unlist(pred_GGM_w.inst)  # Flatten the list to a numeric vector
# Create the time series for pred_GGM
pred_GGM_w_ts <- ts(pred_GGM_w_vec, start = start_time_w, frequency = frequency_w)


plot(sales_w_ts, type= "b",xlab="Week", ylab="Weekly Sales",  pch=16, lty=3, cex=0.6)
lines(pred_GGM_w_ts, col = "red", lty = 2)


```

```{r}
# SARMAX refinement
fit.sales_w <- fitted(ggm1_w)  # Predicted values from the GGM model

if (length(fit.sales_w) != length(sales_w_ts)) {
  stop("fit.sales_w and sales_w_ts lengths do not match")
}

```

```{r}
fit.sales_w <- scale(fit.sales_w) # scale regresor to make convergence

sales_w_ts_scaled <- scale(cumsum(sales_w_ts))  # Scale the time series because if not will not reach convergence

sarima_w <- Arima(
  sales_w_ts_scaled, 
  order = c(1, 0, 1), 
  seasonal = list(order = c(0, 0, 1), period = 52), 
  xreg = fit.sales_w # this is the GGM fitted values
)

summary(sarima_w)

```
```{r}
# get fitted values
# Extract the fitted cumulative values from the SARIMA model
fitted_cumulative <- fitted(sarima_w)

# Reverse scaling transformation to get fitted cumulative values in the original scale
scaling_center <- attr(sales_w_ts_scaled, "scaled:center")
scaling_scale <- attr(sales_w_ts_scaled, "scaled:scale")

fitted_cumulative_original <- fitted_cumulative * scaling_scale + scaling_center

# Convert cumulative fitted values to instantaneous values
fitted_instantaneous <- diff(c(fitted_cumulative_original, NA))  # Add NA to align lengths

# Create a time series object for the fitted instantaneous values
fitted_instantaneous_ts <- ts(
  fitted_instantaneous, 
  start = start(sales_w_ts), 
  frequency = frequency(sales_w_ts)
)

```

```{r}
# Plot original instantaneous values vs fitted instantaneous values
plot(sales_w_ts, type = "p", col = "blue", pch = 16,
     main = "Original vs Fitted Instantaneous Values",
     xlab = "Time", ylab = "Instantaneous Values")

# Add the fitted instantaneous values as a line
lines(fitted_instantaneous_ts, col = "red", lwd = 3, lty = 1)

# Add legend
legend("bottomright", legend = c("Original Instantaneous", "Fitted Instantaneous"),
       col = c("blue", "red"), lty = c(NA, 1), pch = c(16, NA), lwd = c(NA, 3))

```

```{r}
# Residuals
# Step 1: Extract residuals from the SARIMA model
resid_w <- residuals(sarima_w)

# Step 2: Visualize residuals
# Time series plot of residuals
tsdisplay(resid_w, main = "Residual Diagnostics for SARIMA Model")

# Step 3: Test residuals for stationarity
adf_test <- adf.test(resid_w)
cat("ADF Test p-value:", adf_test$p.value, "\n")

if (adf_test$p.value < 0.05) {
  cat("The residuals are stationary.\n")
} else {
  cat("The residuals are not stationary.\n")
}

# Step 4: Test residuals for white noise (no autocorrelation)

ljung_box_test <- Box.test(resid_w, lag = 20, type = "Ljung-Box")
cat("Ljung-Box Test p-value:", ljung_box_test$p.value, "\n")

if (ljung_box_test$p.value > 0.05) {
  cat("The residuals resemble white noise (uncorrelated).\n")
} else {
  cat("The residuals show significant autocorrelation.\n")
}

```

Stationary residuals but with significant correlation

```{r}
#### RMSE for SARIMAX Predictions ####
rmse_mixture_w <- calculate_rmse(observed = sales_w_ts, predicted = fitted_instantaneous_ts)

# Print RMSE for SARIMAX
cat("RMSE for SARIMAX Predictions:", rmse_mixture_w, "\n")
```
#### Daily

```{r}
# Scaling the sales data
sales_min <- min(sales_d_ts)
sales_max <- max(sales_d_ts)
sales_scaled <- (sales_d_ts - sales_min) / (sales_max - sales_min)

# View scaled data
summary(sales_scaled)
plot(sales_scaled, type = "l", main = "Scaled Daily Sales", xlab = "Day", ylab = "Scaled Sales")

#### GGM-------------------------------
# Fit GGM model using scaled data
ggm1_d <- GGM(sales_scaled, mt = 'base', display = T)
summary(ggm1_d)

# Predictions using GGM
pred_GGM_d <- predict(ggm1_d, newx = matrix(1:length(sales_scaled), ncol = 1))
pred_GGM_d.inst <- make.instantaneous(pred_GGM_d)

# Convert predictions to a time series
start_time_d <- start(sales_scaled)  # Start time from scaled sales
frequency_d <- frequency(sales_scaled)  # Frequency from scaled sales

pred_GGM_d_vec <- unlist(pred_GGM_d.inst)  # Convert predictions to a numeric vector
pred_GGM_d_ts <- ts(pred_GGM_d_vec, start = start_time_d, frequency = frequency_d)

# Plot scaled GGM predictions
plot(sales_scaled, type = "b", xlab = "Day", ylab = "Scaled Daily Sales", pch = 16, lty = 3, cex = 0.6)
lines(pred_GGM_d_ts, col = "red", lty = 2)

#### SARIMAX Refinement------------------------

# Use instantaneous fitted values from the GGM model
fit.sales_d_instantaneous <- pred_GGM_d.inst

# Ensure lengths match
if (length(fit.sales_d_instantaneous) != length(sales_scaled)) {
  stop("Instantaneous fitted values and scaled sales data lengths do not match!")
}

# Fit SARIMAX with instantaneous GGM fitted values as regressors
sarima_d <- auto.arima(
  sales_scaled,
  seasonal = TRUE,                 # Enable seasonal components
  xreg = fit.sales_d_instantaneous, # Use instantaneous GGM values as regressors
  stepwise = TRUE,                 # Enable stepwise selection (faster)
  approximation = FALSE            # Use exact maximum likelihood
)

summary(sarima_d)

# Extract fitted scaled values from the SARIMAX model
fitted_scaled <- fitted(sarima_d)

# Reverse scaling for final fitted instantaneous values
fitted_instantaneous_ts <- fitted_scaled * (sales_max - sales_min) + sales_min

# Reverse scaling for GGM predictions
pred_GGM_d_original <- pred_GGM_d_ts * (sales_max - sales_min) + sales_min

# Plot actual vs fitted instantaneous values
plot(sales_d_ts, type = "p", col = "blue", pch = 16,
     main = "Original vs Fitted Instantaneous Values (Daily)",
     xlab = "Time", ylab = "Instantaneous Values")
lines(fitted_instantaneous_ts, col = "red", lwd = 3, lty = 1)

# Add legend
legend("topright", legend = c("Original Instantaneous", "Fitted Instantaneous"),
       col = c("blue", "red"), lty = c(NA, 1), pch = c(16, NA), lwd = c(NA, 3))

```

```{r}
#### Residuals-----------------------
# Extract residuals from the SARIMA model
resid_d <- residuals(sarima_d)

# Visualize residuals
tsdisplay(resid_d, main = "Residual Diagnostics for SARIMA Model")

# Test residuals for stationarity
adf_test <- adf.test(resid_d)
cat("ADF Test p-value:", adf_test$p.value, "\n")

if (adf_test$p.value < 0.05) {
  cat("The residuals are stationary.\n")
} else {
  cat("The residuals are not stationary.\n")
}

# Test residuals for white noise (no autocorrelation)
ljung_box_test <- Box.test(resid_d, lag = 20, type = "Ljung-Box")
cat("Ljung-Box Test p-value:", ljung_box_test$p.value, "\n")

if (ljung_box_test$p.value > 0.05) {
  cat("The residuals resemble white noise (uncorrelated).\n")
} else {
  cat("The residuals show significant autocorrelation.\n")
}

```

```{r}
#### RMSE for SARIMAX Predictions ####
rmse_mixture_d <- calculate_rmse(observed = sales_d_ts, predicted = fitted_instantaneous_ts)

# Print RMSE for SARIMAX
cat("RMSE for SARIMAX Predictions:", rmse_mixture_d, "\n")
```


### Prophet

This model was introduced by Facebook ([S. J. Taylor & Letham,
2018](https://otexts.com/fpp3/prophet.html#ref-prophet)), originally for
forecasting daily data with weekly and yearly seasonality, plus holiday
effects. It was later extended to cover more types of seasonal data. It
works best with time series that have strong seasonality and several
seasons of historical data.

Prophet can be considered a nonlinear regression model (Chapter
[7](https://otexts.com/fpp3/regression.html#regression)), of the form
yt=g(t)+s(t)+h(t)+εt, where g(t) describes a piecewise-linear trend (or
"growth term"), s(t) describes the various seasonal patterns, h(t)
captures the holiday effects, and εt is a white noise error term.

-   The knots (or changepoints) for the piecewise-linear trend are
    automatically selected if not explicitly specified. Optionally, a
    logistic function can be used to set an upper bound on the trend.

-   The seasonal component consists of Fourier terms of the relevant
    periods. By default, order 10 is used for annual seasonality and
    order 3 is used for weekly seasonality.

-   Holiday effects are added as simple dummy variables.

-   The model is estimated using a Bayesian approach to allow for
    automatic selection of the changepoints and other model
    characteristics.

```{r}
library(prophet)
```

The input to Prophet is [always]{.underline} a dataframe with two
columns: ds and y . **The ds (datestamp) column should be of a format,
ideally YYYY-MM-DD for a date or YYYY-MM-DD HH:MM:SS for a timestamp**.
The y column [must]{.underline} be numeric, and represents the
measurement we wish to forecast.

#### Monthly sales

```{r}
# sales montly
ggplot(df_merged_m, aes(x=month, y=sales_m)) +
  geom_line() + ggtitle("Monthly Sales of Restaurant")

head(df_merged_m)
```

```{r}
#Prophet model
# model with no seasonality
df_prophet_m <- df_merged_m[1:2]
head(df_prophet_m)
colnames(df_prophet_m) = c("ds", "y")
df_prophet_m$y <- exp(df_prophet_m$y)
prophet_sales_m <- prophet(df_prophet_m)
```

```{r}
head(df_prophet_m)
```

```{r}
# Step 2: Create a future dataframe for the next 14 months
future_sales_m <- make_future_dataframe(
  prophet_sales_m,
  periods = 14,           # Forecast for 14 months
  freq = 'month',         # Monthly frequency
  include_history = TRUE  # Include historical data in the future dataframe
)
tail(future_sales_m)
```

```{r}
forecast_sales_m <- predict(prophet_sales_m, future_sales_m)
tail(forecast_sales_m[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])

plot(prophet_sales_m, forecast_sales_m)
```

```{r}
prophet_plot_components(prophet_sales_m, forecast_sales_m)
```

```{r}
dyplot.prophet(prophet_sales_m, forecast_sales_m)
```

```{r}
#Use the original dataframe to get the fitted values
fitted_values <- predict(prophet_sales_m, df_prophet_m)

# Extract the fitted values (column 'yhat' contains the fitted values)
fitted_y <- fitted_values$yhat

# Calculate RMSE

actual_y <- df_prophet_m$y  # Actual sales values
rmse_prophet_m <- calculate_rmse(observed = actual_y, predicted = fitted_y)

# Print RMSE
cat("RMSE for Prophet Fitted Values:", rmse_prophet_m, "\n")
```
Residuals for prophet
```{r}
# Calculate Residuals
residuals_prophet <- actual_y - fitted_y  # Residuals = Actual - Fitted

#  Visualize Residuals using tsdisplay

tsdisplay(residuals_prophet, main = "Residual Diagnostics for Prophet Model")

#  Perform ADF Test for Stationarity

adf_test <- adf.test(residuals_prophet)
cat("ADF Test p-value:", adf_test$p.value, "\n")

if (adf_test$p.value < 0.05) {
  cat("Residuals are stationary (reject H0).\n")
} else {
  cat("Residuals are not stationary (fail to reject H0).\n")
}

#  Perform Serial Correlation Test
ljung_box_test <- Box.test(residuals_prophet, lag = 10, type = "Ljung-Box")
cat("Ljung-Box Test p-value:", ljung_box_test$p.value, "\n")

if (ljung_box_test$p.value > 0.05) {
  cat("Residuals resemble white noise (no significant autocorrelation).\n")
} else {
  cat("Residuals show significant autocorrelation.\n")
}

```


#### Weekly sales

```{r}
ggplot(df_merged_w, aes(x=week, y=sales_w)) +
  geom_line() + ggtitle("Weekly Sales of Restaurant")
```

```{r}
head(df_merged_w)
```

```{r}
#Prophet model
# model with no seasonality
df_prophet_w <- df_merged_w[1:2]
colnames(df_prophet_w) = c("ds", "y")
df_prophet_w$y <- exp(df_prophet_w$y)
df_prophet_w


prophet_sales_w <- prophet(df_prophet_w)
```

Predictions are made on a dataframe with a column `ds` containing the
dates for which predictions are to be made. The `make_future_dataframe`
function takes the model object and a number of periods to forecast and
produces a suitable dataframe. By default it will also include the
historical dates so we can evaluate in-sample fit.

```{r}
future_sales_w <- make_future_dataframe(prophet_sales_w, 
                                        periods = 52,
                                        freq = 'week',
                                        include_history = T)
tail(future_sales_w)
```

As with most modeling procedures in R, we use the generic `predict`
function to get our forecast. The `forecast` object is a dataframe with
a column `yhat` containing the forecast. It has additional columns for
uncertainty intervals and seasonal components.

```{r}
# R
forecast_sales_w <- predict(prophet_sales_w, future_sales_w)
tail(forecast_sales_w[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])
```

```{r}
plot(prophet_sales_w, forecast_sales_w)
```

You can use the `prophet_plot_components` function to see the forecast
broken down into trend, weekly seasonality, and yearly seasonality.

```{r}
prophet_plot_components(prophet_sales_w, forecast_sales_w)
```

```{r}
dyplot.prophet(prophet_sales_w, forecast_sales_w)
```

```{r}
# Use the original dataset to get fitted values
fitted_values_w <- predict(prophet_sales_w, df_prophet_w)

# Extract the fitted values (column 'yhat' contains the fitted values)
fitted_y_w <- fitted_values_w$yhat

# Ensure alignment between actual values (y) and fitted values (yhat)
actual_y_w <- df_prophet_w$y  # Actual weekly sales values

# Calculate RMSE for weekly data
rmse_prophet_w <- calculate_rmse(observed = actual_y_w, predicted = fitted_y_w)

# Print RMSE
cat("RMSE for Prophet Fitted Values (Weekly):", rmse_prophet_w, "\n")
```

```{r}
# Calculate Residuals
residuals_prophet_w <- actual_y_w - fitted_y_w  # Residuals = Actual - Fitted

# Visualize Residuals using tsdisplay
tsdisplay(residuals_prophet_w, main = "Residual Diagnostics for Weekly Prophet Model")

# Perform ADF Test for Stationarity

adf_test_w <- adf.test(residuals_prophet_w)
cat("ADF Test p-value:", adf_test_w$p.value, "\n")

if (adf_test_w$p.value < 0.05) {
  cat("Residuals are stationary (reject H0).\n")
} else {
  cat("Residuals are not stationary (fail to reject H0).\n")
}

# Perform Serial Correlation Test
ljung_box_test_w <- Box.test(residuals_prophet_w, lag = 10, type = "Ljung-Box")
cat("Ljung-Box Test p-value:", ljung_box_test_w$p.value, "\n")

if (ljung_box_test_w$p.value > 0.05) {
  cat("Residuals resemble white noise (no significant autocorrelation).\n")
} else {
  cat("Residuals show significant autocorrelation.\n")
}

```

#### Daily Sales

```{r}
head(sales_d_ts)

plot(sales_d_ts)
```

```{r}
sales_d_values <- as.numeric(sales_d_ts)   # Extract numeric values

df_prophet_d <- data.frame(
  ds = df_merged_d$date,  # Dates
  y = sales_d_values   # Sales values
)

```

```{r}
#Prophet model


#prophet_sales_d <- prophet(df_prophet, weekly.seasonality = TRUE)
prophet_sales_d <- prophet(df_prophet_d)
```

```{r}
future_sales_d <- make_future_dataframe(prophet_sales_d,
                                        periods = 60,
                                        freq = 'day',
                                        include_history = T)
tail(future_sales_d)
```

```{r}
forecast_sales_d <- predict(prophet_sales_d, future_sales_d)
tail(forecast_sales_d[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])
```

```{r}
plot(prophet_sales_d, forecast_sales_d)
```

```{r}
prophet_plot_components(prophet_sales_d, forecast_sales_d)
```

```{r}
dyplot.prophet(prophet_sales_d, forecast_sales_d)
```


```{r}
# Extract fitted values for RMSE calculation
fitted_values_d <- predict(prophet_sales_d, df_prophet_d)

# Extract fitted values (column 'yhat')
fitted_y_d <- fitted_values_d$yhat
actual_y_d <- df_prophet_d$y  # Actual sales values

# Step 8: Calculate RMSE
rmse_prophet_d <- calculate_rmse(observed = actual_y_d, predicted = fitted_y_d)

# Print RMSE
cat("RMSE for Prophet Fitted Values (Daily):", rmse_prophet_d, "\n")
```


```{r}
# Calculate Residuals
residuals_prophet_d <- actual_y_d - fitted_y_d  # Residuals = Actual - Fitted

# Visualize Residuals using tsdisplay

tsdisplay(residuals_prophet_d, main = "Residual Diagnostics for Daily Prophet Model")

# Perform ADF Test for Stationarity

adf_test_d <- adf.test(residuals_prophet_d)
cat("ADF Test p-value:", adf_test_d$p.value, "\n")

if (adf_test_d$p.value < 0.05) {
  cat("Residuals are stationary (reject H0).\n")
} else {
  cat("Residuals are not stationary (fail to reject H0).\n")
}

# Perform Serial Correlation Test
ljung_box_test_d <- Box.test(residuals_prophet_d, lag = 20, type = "Ljung-Box")
cat("Ljung-Box Test p-value:", ljung_box_test_d$p.value, "\n")

if (ljung_box_test_d$p.value > 0.05) {
  cat("Residuals resemble white noise (no significant autocorrelation).\n")
} else {
  cat("Residuals show significant autocorrelation.\n")
}

```



```{r}
rmse_list <- c(rmse_ols_m, rmse_ols_w, rmse_ols_d,
               rmse_bm_m, rmse_bm_w, rmse_bm_d,
               rmse_ggm1, rmse_ggm_w, rmse_ggm_d,
               rmse_hw2,
               rmse_auto_arima, rmse_auto_arima_w, rmse_auto_arima_d,
               rmse_sarima_d,
               rmse_sarimax_d,
               rmse_exp_sm_m, rmse_exp_sm_w, rmse_exp_sm_m,
               rmse_mixture_m, rmse_mixture_w, rmse_mixture_d,
               rmse_prophet_m, rmse_prophet_w, rmse_prophet_d
               )
rmse_list
```
## Evaluation of all models
```{r}
# Initialize an empty data frame for RMSE values
rmse_table <- data.frame(
  Model = character(),
  Monthly = numeric(),
  Weekly = numeric(),
  Daily = numeric(),
  stringsAsFactors = FALSE
)

# Monthly RMSE values
rmse_monthly <- c(
  "OLS" = rmse_ols_m,
  "Bass_Model" = rmse_bm_m,
  "GGM" = rmse_ggm1,
  "Holt_Winters" = rmse_hw2,
  "Arima" = rmse_auto_arima,
  "Exp_Smooth" = rmse_exp_sm_m,
  "GGM+SARIMA" = rmse_mixture_m,
  "Prophet" = rmse_prophet_m
)

# Weekly RMSE values
rmse_weekly <- c(
  "OLS" = rmse_ols_w,
  "Bass_Model" = rmse_bm_w,
  "GGM" = rmse_ggm_w,
  "Holt_Winters" = NaN,
  "Arima" = rmse_auto_arima_w,
  "Exp_Smooth" = rmse_exp_sm_w,
  "GGM+SARIMA" = rmse_mixture_w,
  "Prophet" = rmse_prophet_w
)

# Daily RMSE values
rmse_daily <- c(
  "OLS" = rmse_ols_d,
  "Bass_Model" = rmse_bm_d,
  "GGM" = rmse_ggm_d,
  "Holt_Winters" = NaN,
  "Arima" = rmse_auto_arima_d,
  "Exp_Smooth" = rmse_exp_sm_d,
  "GGM+SARIMA" = rmse_mixture_d,
  "Prophet" = rmse_prophet_d
)

# Combine RMSE values into a table
for (model_name in names(rmse_monthly)) {
  rmse_table <- rbind(rmse_table, data.frame(
    Model = model_name,
    Monthly = rmse_monthly[model_name],
    Weekly = rmse_weekly[model_name],
    Daily = rmse_daily[model_name]
  ))
}

# View the RMSE table
print(rmse_table)

```

Best models are:

* Monthly: GGM + SARIMA
* Weekly: GGM + SARIMA
* Daily: Prophet

## Evaluation of best models on Test Set
### Import test set
```{r}
# target variable
test_sales_df <- read_excel("data/sales/test_data.xlsx")
head(test_sales_df)
```
```{r}
df_sales_m_test <- test_sales_df %>%
  mutate(month = floor_date(date, "month")) %>% # Extract month
  group_by(month) %>%
  summarise(sales_m = sum(sales_cop), bar_m = sum(bar), food_m = sum(food)
            )     # Summing values

head(df_sales_m_test)
```
```{r}
## sales weekly
df_sales_w_test <- test_sales_df %>%
  mutate(week = floor_date(date, "week")) %>% # Extract month
  group_by(week) %>%
  summarise(sales_w = sum(sales_cop), bar_w = sum(bar), food_w = sum(food))     # Summing values

head(df_sales_w_test)
```
## Forecast vs Actual
### Montly
```{r}
```
```{r}
cumsum(sales_m_ts)
forecast_cumulative
```

### Weekly
### Daily

## Forecast with Best Models
### Montly
### Weekly
### Daily
